Multilevel computation of glacier geometry \\ from Stokes dynamics

### 2021-06-13 paper/msg.tex
Attempt to put ideas in some order.

In this section we propose solvers for \eqref{eq:fesigpweakform}, namely certain projected and nonlinear iterations which are variants of the Gauss-Seidel and Jacobi methods, but made suitable for VIs \cite{KinderlehrerStampacchia1980}.  Given a current surface elevation $s^h\in \mathcal{K}^h$, these iterations sweep through the $m$ interior vertices ($P_1$ nodes) of $\mathcal{T}$ in a fixed order, solving one-dimensional VI problems.  While these iterations are adequate smoothers in a multilevel method (next section), they converge slowly if used as single-level methods.  Because of the non-locality and density of the IDO $\Phi^h$ (Section \ref{sec:weakido}), we must make careful approximations in order to retain adequate per-iteration performance.

Denote an arbitrary vertex of $\mathcal{T}$ by $(x_j,y_j)$ and let $\psi_j \in \mathcal{V}^h$ be the corresponding $P_1$ hat function, with $\psi_j(x_k,y_k)=\delta_{jk}$ \cite{Elmanetal2014}.  Let $\mathcal{W}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0 \text{ and } t^h|_{\partial\Omega} = 0\}$.  Let $i=0,\dots,m-1$ index the interior vertices of $\mathcal{T}$, so $(x_i,y_i) \notin \partial\Omega$.  Note that $\psi_i \in \mathcal{W}_+^h$, and $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.  We call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009} for $s^h$.

The one-dimensional VI at interior node $i$ is derived from the discrete SIGP \eqref{eq:fesigpweakform} using $r = s^h+\gamma \psi_i$ and $s = s^h+c \psi_i$ for some scalars $\gamma,c$.  Specifically, we seek $c \ge \beta_i^{s^h}$ such that
\begin{equation}
F^h(s^h+c \psi_i)[(s^h+\gamma \psi_i) - (s^h+c \psi_i)] \ge \ip{a}{(s^h+\gamma \psi_i) - (s^h+c \psi_i)} \label{eq:fepointwiseviEARLY}
\end{equation}
for all $\gamma \ge \beta_i^{s^h}$.  Let
\begin{equation}
\rho_i(s^h; c) = F^h(s^h+c\psi_i)[\psi_i] - \ip{a}{\psi_i} \label{eq:ferhoi}
\end{equation}
be the point-wise residual function, and observe that \eqref{eq:fepointwiseviEARLY} simplifies to
\begin{equation}
(\gamma - c) \,\rho_i(s^h; c) \ge 0. \label{eq:fepointwisevi}
\end{equation}
The smoother will approximately solve \eqref{eq:fepointwisevi} for $c$ the smoother, and then update $s^h \gets s^h + c \psi_i$.  Observe that admissibility is preserved because $s^h+c \psi_i \in \mathcal{K}^h$ when $c \ge \beta_i^{s^h}$.

If $\rho_i(s^h; c)$ were linear and increasing in $c$, e.g.~$\rho_i(s^h; c) = \rho_i(s^h; 0) + \alpha c$ with $\alpha > 0$, then the solution to \eqref{eq:fepointwisevi} would be given by the simple formula $c = \max\left\{-\rho_i(s^h; 0)/\alpha, \beta_i^{s^h}\right\}$ \cite{GraeserKornhuber2009}.  In fact $\rho_i(s^h; c)$ is nonlinear, and we will linearize
\begin{equation}
\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, \rho_i'(s^h; 0), \label{eq:rhoapprox}
\end{equation}
and at each node where $\rho_i'(s^h; 0) > 0$ we will do one projected Newton step using the simple formula.  If $\rho_i'(s^h; 0) \le 0$, so that the model has degenerated in the sense that the linearized, pointwise operator is not acting elliptically with respect to a perturbation, we set $c = \beta_i^{s^h}$ to remove the ice at that location.  % FIXME TOO AGGRESSIVE?

Because the interior PDE in the shallow, flat-bed model is elliptic, it follows that $\rho_i'(s^h; 0) > 0$ on the icy part of $\Omega$ \cite{JouvetBueler2012}.  However, for the full Glen-Stokes model there is no such guarantee to our knowledge.

Note that the derivative $\rho_i'(s^h; 0)$, which is also a diagonal entry of the Jacobian of $F^h$, can be approximated by a finite difference,
\begin{equation}
\rho_i'(s^h; 0) = (F^h)'(s^h)[\psi_i,\psi_i] \approx \frac{\rho_i(s^h; \eps) - \rho_i(s^h; 0)}{\eps}.  \label{eq:rhofd}
\end{equation}
Literal use of \eqref{eq:rhofd} is very expensive because each value $\rho_i(s^h; \eps)$ requires a separate Glen-Stokes computation for the residual.

If each pointwise update $s^h \gets s^h + c \psi_i$ is done immediately then the smoother is of Gauss-Seidel (GS) type, also called serial or multiplicative, while a (parallel, additive) Jacobi smoother computes and stores solutions $c$ at every node $i$ before doing any updates.  In linear problems GS generally converges in about half as many iterations \cite{Greenbaum1997}, but here a more glaring difference dominates, namely that GS imposes the expense of yet more residual evaluations.  Nonetheless, the GS iteration using a single Newton step and approximation \eqref{eq:rhofd} is a well-defined smoother, so we state it in Pseudocode \ref{pc:pngsslow}.  It is given in a form which also allows over-relaxation ($\id{omega} > 1$) or under-relaxation ($\id{omega} < 1$) if desired.  Note that the Glen-Stokes problem \eqref{eq:glenstokesweak} is solved $2m$ times per application of this function.

\begin{pcode}[ht]
\begin{pseudo*}
\pr{pngs\_slow}(s^h,b^h,\id{eps}=1.0,\id{omega}=1.0)\text{:} \\+
    \ct{check admissibility: $s^h \ge b^h$} \\
    for $i = 0,\dots,m-1$ \\+
        $\alpha_i = (\rho_i(s^h; \eps) - \rho_i(s^h; 0))/\eps$  \qquad\qquad \ct{Jacobian diagonal entry} \\
        if $\alpha_i > 0$ \\+
            $c_i = - \rho_i(s^h; 0) / \alpha_i$ \\
            $(s^h)_i \gets \max\{(s^h)_i + \id{omega}\,c_i, (b^h)_i\}$ \\-
        else \\+
            $(s^h)_i \gets \beta_i$ \qquad\qquad \ct{non-elliptic case}
\end{pseudo*}
\caption{Projected nonlinear GS iteration, an expensive in-place smoother using a finite-difference (FD) derivative for the Jacobian diagonal.}
\label{pc:pngsslow}
\end{pcode}

@@

FIXME SMOOTHER POSSIBILITIES.  1 COLORING (PSEUDO-COLORING WITH LATERAL SEPARATION OF A FEW ICE THICKNESSES).  2 PATCH-TYPE WHICH SOLVE LOW-DIMENSIONAL VI.

@@

>>>>

In this section we consider single-level solvers for \eqref{eq:fesigpweakform}, namely certain nonlinear iterations which are made suitable for VIs \cite{KinderlehrerStampacchia1980}.  While these iterations are, in some cases, adequate smoothers in a multilevel method (next section), they converge slowly if used directly as solvers.  Because of the non-locality and density of the IDO $\Phi^h$ (Section \ref{sec:weakido}), we must make careful approximations in order to retain adequate per-iteration performance.  These iterative solvers, \emph{smoothers} from now on, can be categorized by the number of nodes which are adjusted at each iteration: pointwise, patch-type, and global.

Suppose there are $m$ interior vertices ($P_1$ nodes) of $\mathcal{T}$, denote a vertex in $\mathcal{T}$ by $(x_j,y_j)$, and let $\psi_j \in \mathcal{V}^h$ be the corresponding $P_1$ hat function, with $\psi_j(x_k,y_k)=\delta_{jk}$.  Let $s^h\in \mathcal{K}^h$ be the current surface elevation iterate.  The Richardson pointwise smoother sweeps through the nodes and subtracts-off the residual: $s^h \gets s^h - \alpha (F^h(s^h)-a)$ where $\alpha>0$ must be chosen so that the iteration converges.  By contrast the Gauss-Seidel, and Jacobi pointwise smoothers sweep through the nodes solving one-dimensional VI problems for a value $c\in \RR$ (below).  If each pointwise update $s^h \gets s^h + c \psi_i$ is done immediately then the smoother is of Gauss-Seidel (GS) type, also called serial or multiplicative, while a (parallel, additive) Jacobi smoother computes and stores solutions $c$ at every node before doing any updates.  In linear problems GS generally converges in about half as many iterations \cite{Greenbaum1997}.

To be more precise, let $\mathcal{W}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0 \text{ and } t^h|_{\partial\Omega} = 0\}$.  If $i=0,\dots,m-1$ index the interior vertices of $\mathcal{T}$, note that $\psi_i \in \mathcal{W}_+^h$, and that $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.  (One might call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009} for $s^h$.)  The one-dimensional VI at interior node $i$ is derived from the discrete SIGP \eqref{eq:fesigpweakform} by replacements $s^h \to s^h+c \psi_i$, $r^h \to r^h + \gamma \psi_i$ for some scalars $c,\gamma$.  We then seek $c \ge \beta_i^{s^h}$ such that
\begin{equation}
F^h(s^h+c \psi_i)[(s^h+\gamma \psi_i) - (s^h+c \psi_i)] \ge \ip{a}{(s^h+\gamma \psi_i) - (s^h+c \psi_i)} \label{eq:fepointwiseviEARLY}
\end{equation}
for all $\gamma \ge \beta_i^{s^h}$.  Let
\begin{equation}
\rho_i(s^h; c) = F^h(s^h+c\psi_i)[\psi_i] - \ip{a}{\psi_i} \label{eq:ferhoi}
\end{equation}
be the pointwise residual function, and observe that \eqref{eq:fepointwiseviEARLY} simplifies to
\begin{equation}
(\gamma - c) \,\rho_i(s^h; c) \ge 0. \label{eq:fepointwisevi}
\end{equation}
The smoother approximately solves \eqref{eq:fepointwisevi} for $c$ and then updates $s^h \gets s^h + c \psi_i$ either in serial or parallel.  Observe that admissibility is preserved because $s^h+c \psi_i \in \mathcal{K}^h$ when $c \ge \beta_i^{s^h}$.

If $\rho_i(s^h; c)$ were linear and increasing in $c$, e.g.~$\rho_i(s^h; c) = \rho_i(s^h; 0) + \alpha c$ with $\alpha > 0$, then the solution to VI \eqref{eq:fepointwisevi} would be given by the simple formula $c = \max\left\{-\rho_i(s^h; 0)/\alpha, \beta_i^{s^h}\right\}$ \cite{GraeserKornhuber2009}.  In fact $\rho_i(s^h; c)$ is nonlinear, and we will linearize
\begin{equation}
\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, \rho_i'(s^h; 0). \label{eq:rhoapprox}
\end{equation}
At each node where $\rho_i'(s^h; 0) > 0$ we will do one projected Newton step using the simple formula.  However, if $\rho_i'(s^h; 0) \le 0$, so that the model has degenerated in the sense that the linearized, pointwise operator is not acting elliptically with respect to a perturbation, we set $c = \beta_i^{s^h}$ to remove the ice at that location.  % FIXME TOO AGGRESSIVE?

Using a single Newton step and approximation \eqref{eq:rhofd}, GS is a well-defined smoother, so we state it in Pseudocode \ref{pc:pngsslow}.  It is given in a form which also allows over-relaxation ($\id{omega} > 1$) or under-relaxation ($\id{omega} < 1$) if desired.  Note that the Glen-Stokes problem \eqref{eq:glenstokesweak} is solved $2m$ times per application of this function.

\begin{pcode}[ht]
\begin{pseudo*}
\pr{pngs}(s^h,b^h,\id{omega}=1.0)\text{:} \\+
    \ct{check admissibility: $s^h \ge b^h$} \\
    for $i = 0,\dots,m-1$ \\+
        $\alpha_i = \rho_i'(s^h; 0)$  \qquad\qquad \ct{Jacobian diagonal entry} \\
        if $\alpha_i > 0$ \\+
            $c_i = - \rho_i(s^h; 0) / \alpha_i$ \\
            $(s^h)_i \gets \max\{(s^h)_i + \id{omega}\,c_i, (b^h)_i\}$ \\-
        else \\+
            $(s^h)_i \gets \beta_i$ \qquad\qquad \ct{non-elliptic case}
\end{pseudo*}
\caption{Projected nonlinear GS iteration, a conceptual in-place, pointwise smoother which solves one-dimensional VIs \eqref{eq:fepointwisevi} in serial.}
\label{pc:pngsslow}
\end{pcode}

Patch-type smoothers \cite{Farrelletal2019} have to our knowledge not been used for VI problems before, but they would sweep through the nodes solving local VIs of small dimension $\ell>1$, and again there would be serial and parallel versions.  However, we first consider the opposite extreme from a pointwise smoother, namely a global smoother in which all the degrees of freedom are coupled together into an algebraic system.

FIXME reduced-space Newton solver using a banded Jacobian approximation

\section{Approximate Jacobians} \label{sec:jacobians}

The above pseudocodes assume the accessibility of the diagonal of the Jacobian (pointwise smoothers) or of arbitrary entries (Newton solver).

Because the interior PDE in the SIA is elliptic, it follows that $\rho_i'(s^h; 0) > 0$ on the icy part of $\Omega$ \cite{JouvetBueler2012}, but for the full Glen-Stokes model there is no such guarantee to our knowledge.  Note that the derivative $\rho_i'(s^h; 0)$, which is also a diagonal entry of the Jacobian of $F^h$, can be approximated by a finite difference,
\begin{equation}
\rho_i'(s^h; 0) = (F^h)'(s^h)[\psi_i,\psi_i] \approx \frac{\rho_i(s^h; \eps) - \rho_i(s^h; 0)}{\eps}.  \label{eq:rhofd}
\end{equation}
Literal use of \eqref{eq:rhofd} is very expensive because each value $\rho_i(s^h; \eps)$ requires a separate Glen-Stokes computation for the residual.

A glaring difference between the GS and Jacobi smoothers now dominates.  Namely, GS imposes the expense of far more residual evaluations.

FIXME finite-difference approximation using pseudo-coloring based on ice thicknesses or rather longitudinal coupling length \cite{KambEchelmeyer1986}

@@

FIXME PSEUDO-COLORING WITH LATERAL SEPARATION OF A FEW ICE THICKNESSES

@@

\section{Performance models} \label{sec:perfmodels}

FIXME
